---
title: "Asthma data: Preliminary Data Exploration 3 - Graphical Summary"
output:
  html_document:
    toc: true
    toc_depth: 4
---

## Intro

[previous section](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/02_Exploratory_Analysis/asthma-prelim_data_exploration-02.md).

This document is a first look at the previously processed asthma dataset.

Playinng around with ideas... State level (50 states) granularity needs different strategy to condense exploratory vis over time. a) raw numbers b) relative to value in current year (0) previous years are either higher or lower!.
What vis types?

### Setup

In the [data preparation phase](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/01_Data_Prep/asthma_data_source_prep_02.Rmd), we were able to scrape together a number of tables from the web and save them to disk as CSV files. 


```{r, configChunk, echo=TRUE}
# Getting knitr to play ball
# http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
require(knitr, quietly=TRUE)
opts_knit$set(root.dir = normalizePath("../"))
```


## Data exploration

### Data import

We will import the processed Asthma dataset as per the [previous section](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/02_Exploratory_Analysis/asthma-prelim_data_exploration-02.Rmd), but with helper functions necessary to facilitate our graphical data analysis.

```{r, dataImport, message=F, warning=F}
#### 1. Data Import ####
# import helper functions
helperParentDir = "./helper_scripts/"

library(ggplot2); library(grid); library(gridExtra)

source(file.path(helperParentDir, "01_data_prep/", "general_helpers_01.R"))
source(file.path(helperParentDir, "02_data_explore/", "helpers-asthma_data_explore_plots_01.R"))

# Source data dir
asthmaDataDir = "./Data/asthma/series/"
asthmaGZipFile = file.path(asthmaDataDir, "asthma_data.csv.gz")

# readr handles compressed files automatically :)
asthmaData = gen_helpers_01$read_csv_quiet(asthmaGZipFile)
# but need to convert to data.frame ... tibble issues!
asthmaData = as.data.frame(asthmaData)
```

### Initial data snapshot

Then: try to plot data:

Groups (2): Adult and Child, by Demographic (6) across Time (15 years)

State level (50 states) granularity needs different strategy to condense exploratory vis over time. a) raw numbers b) relative to value in current year (0) previous years are either higher or lower!.
What vis types?

.......
```{r, createExploreGraphs}
asthmaExplorePlots = data_exp_asthma_01$asthaPlotExploreGroup(asthmaData)
sort(names(asthmaExplorePlots))
```

### Graphs by Demographic Category

#### Age

```{r, ageExploreGraphs}
ageGraphList = grepl("age", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[ageGraphList], ncol = 1))
```

#### Ethnicity

```{r, ethnicExploreGraphs}
ethnicGraphList = grepl("ethnicity", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[ethnicGraphList], ncol = 1))
```

#### Gender

```{r, genderExploreGraphs}
genderGraphList = grepl("gender", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[genderGraphList], ncol = 1))
```

#### Income

```{r, incomeExploreGraphs}
incomeGraphList = grepl("income", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[incomeGraphList], ncol = 1))
```

#### Education

```{r, educationExploreGraphs}
educationGraphList = grepl("education", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[educationGraphList], ncol = 1))
```

#### Overall

```{r, overallExploreGraphs}
overallGraphList = grepl("overall", names(asthmaExplorePlots))
grid.draw(arrangeGrob(grobs = asthmaExplorePlots[overallGraphList], ncol = 1))
```

[next section](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/02_Exploratory_Analysis/asthma-prelim_data_exploration-04.md).

<p style="text-align:center; border-style:solid;border-color:blue;">:smile: Back to [Project Readme](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/README.md#2-exploratory-data-analysis) :alien: Back to [the top]() :smile:</p><br/><br/>