---
title: "Asthma data: Preliminary Data Exploration 4 - Further analysis"
output:
  html_document:
    toc: true
    toc_depth: 4
    keep_md: true
---

## Intro

This document is a first look at the previously processed asthma dataset.

[previous section](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/02_Exploratory_Analysis/asthma-prelim_data_exploration-03.md).

### Setup


```{r, configChunk, echo=TRUE}
# Getting knitr to play ball
# http://stackoverflow.com/questions/24585254/working-with-knitr-using-subdirectories
require(knitr, quietly=TRUE)
opts_knit$set(root.dir = normalizePath("../"))
```


## Data exploration

### Data import

We will import the processed Asthma dataset as per the [previous section](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/02_Exploratory_Analysis/asthma-prelim_data_exploration-02.Rmd), but with helper functions necessary to facilitate our graphical data analysis.

```{r, dataImport, message=F, warning=F}
#### 1. Data Import ####
# import helper functions
helperParentDir = "./helper_scripts/"

library(ggplot2); library(grid); library(gridExtra)

source(file.path(helperParentDir, "01_data_prep/", "general_helpers_01.R"))
source(file.path(helperParentDir, "02_data_explore/", "helpers-asthma_data_explore_plots_01.R"))

# Source data dir
asthmaDataDir = "./Data/asthma/series/"
asthmaGZipFile = file.path(asthmaDataDir, "asthma_data.csv.gz")

# readr handles compressed files automatically :)
asthmaData = gen_helpers_01$read_csv_quiet(asthmaGZipFile)
# but need to convert to data.frame ... tibble issues!
asthmaData = as.data.frame(asthmaData)
```

### Strategy Outline

Previous analysis gave us a simple overview of the aggregate data at a national level (via the totals).

Trying to measure state-level data integrity, to see if the dataset can be explored at a regional or state level or if the data is too patchy (particularly in the earlier years). This also gives us an idea of whether we can apply data imputation.

### The search for missing data

Lists all the state and demographic categories

| State | Demographic Type Cols                                 |
|-------|-------------------------------------------------------|
| list  | count of missing years for a given demographic column |

Gives a state-level view of the completeness of the data in each region. Captures time by counting the number of years where data is missing for a particular demographic type for each region.

The important thing is to try to see whether it is worth modelling the data by State. Summarising all of the essential characteristics of this dataset

* geographic: State (Rows)
* demographic: Column variables
* temporal: Missing years (table cells)

This allows us to thus:

* decide which data subsets are useful
* identify data demographic subsets worth examining in greater detail
* identify potential data imputation strategies (where relevant)

We can do the same thing by attributes

| Year  | Demographic Type Cols                                 |
|-------|-------------------------------------------------------|
| list  | count of missing years for a given demographic column |

* geographic: State (table cells)
* demographic: Column variables
* temporal: Year (Rows)

Conceptually, this can basically be thought of as a quasi-pivot table


<p style="text-align:center; border-style:solid;border-color:blue;">:smile: Back to [Project Readme](https://github.com/drbulu/healthyAir_DSc_proj/blob/master/README.md#2-exploratory-data-analysis) :alien: Back to [the top]() :smile:</p><br/><br/>